---
title:   "Shen / Liss Model Fitting Report"
date:    "`r paste0('Date: ', format(Sys.time(), '%Y-%m-%d %X %Z'))`"
author: "Prepared by James Robert White, PhD (Resphera Biosciences)" 
output:
  html_document:
    theme: default
    highlight: default
    toc: true
    toc_depth: 3
    collapse: subsubsection
    number_sections: true
    toc_float:
      collapsed: false
      smooth_scroll: true
---
<style type="text/css">
.list-group-item.active {
  color: white;
  background-color: #3a86ff;
  border: none;
}
.list-group-item:hover,
.list-group-item.active:hover {
    color: white;
    background-color: #3a86ff;
}
.title{
  color: #3a86ff;
}
</style>


```{r dirs, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
# --- 
# to run:
# setwd("~/Desktop/shen-model-fitting")
# library(rmarkdown)
# render("A01.1-shen-report.Rmd")

# ---
# otherwise, for troubleshooting continue step by step
# set the working directory as the location of the script
setwd("~/Desktop/shen-model-fitting")
# ---

knitr::opts_chunk$set(echo=FALSE)
library(RColorBrewer)
library(ggplot2)
library(scales)
require(gtools)
library(MASS)
library(reshape2)
library(ggrepel)
library(tidyverse)
library(table1)
library(ggpubr)
library(readxl)
library(DT)
library(pROC)
library(DescTools)
library(caret)
library(glmnet)


set.seed(123)

# ---
# load data
cncrDf = read.csv("Logistic Regression Cancer Cell.csv", check.names=FALSE)
tcgaDf = read.csv("Logistic Regression TCGA.csv", check.names=FALSE)

# ---
# Our primary goal in this analysis is to assist with the
# response to the following reviewer comment:
# "The authors simply averaged the Z-scores of each gene equally
# without any adjusted weighting or attempts to use more 
# sophisticated regression or elastic net methods. The authors 
# may want to test if these approaches improve the performance 
# of their signature on the TCGA and Taylor datasets" 
# ---

# ---
# merge data
cncrDf  = cncrDf[,c(1:9)]
colnames(cncrDf) = c("id", colnames(cncrDf)[2:8], "Gleason Score")
cncrDf$cohort = "Cancer Cell"

tcgaDf  = tcgaDf[,c(1:7,9:10)]
tcgaDf$cohort = "TCGA"
tcgaDf  = tcgaDf[,c(1:6,8:9,7,10)]
colnames(tcgaDf) = colnames(cncrDf) 

jointDf = rbind(cncrDf,tcgaDf)
jointDf = jointDf[,c(1,10,2:9)]

myDT1 <- datatable(jointDf, 
                  class      = 'cell-border stripe', 
                  rownames   = F, 
                  filter     = 'top',
                  extensions = "Buttons", 
                  options    = list(paging = TRUE,
                                    scrollX=TRUE, 
                                    searching = TRUE,
                                    ordering = TRUE,
                                    dom = 'Bfrtip',
                                    buttons = c('copy', 'csv', 'excel'),
                                    pageLength =20))

for (j in 3:7){
  jointDf[,j] = as.numeric(jointDf[,j])
}

for (j in 3:7){
  jointDf = jointDf[!is.na(jointDf[,j]),]
}

jointDf$cohort = factor(jointDf$cohort)
jointDf        = jointDf[grepl("^[01]", jointDf$`Disease Free Status`),]
jointDf$`Disease Free Status` = ifelse(grepl("^0", jointDf$`Disease Free Status`), "disease free", "recurred")
jointDf$`Disease Free Status` = factor(jointDf$`Disease Free Status`)

myDT2 <- datatable(aggregate(id~`Disease Free Status`+cohort, jointDf, FUN=length))

meltedjointDf = melt(jointDf, id.vars=colnames(jointDf)[c(1,2,8,9,10)])
colnames(meltedjointDf)[6] = "gene"
colnames(meltedjointDf)[7] = "Value"

myCol1 = c("#3a86ff","#ffbe0b")

plotDist <- ggplot(data=meltedjointDf, aes_string(x=paste0("`", "Disease Free Status","`"), y=paste0("`", "Value","`"))) +
  geom_boxplot(aes(color=`Disease Free Status`), fill=NA, outlier.size=0, coef=1e100, alpha=0.9) +
  geom_point(aes(color=`Disease Free Status`), pch=19, size=1.5, alpha=0.35)  +
  scale_fill_manual(values=myCol1) + 
  scale_color_manual(values=myCol1) + 
  theme_bw() + 
  theme(aspect.ratio = 0.75, 
        legend.position = "right", 
        legend.text  = element_text(size=8), 
        legend.title = element_text(size=8),
        axis.text.x  = element_text(size=10, colour="black"),
        axis.text.y  = element_text(size=10, colour="black"),
        axis.title.x = element_text(size=11, colour="black"),
        axis.title.y = element_text(size=11, colour="black"),
        plot.title   = element_text(size=10, colour="black"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) + 
  scale_y_continuous(breaks=pretty_breaks(n=5), expand=expansion(mult = c(0.05, 0.15))) +
  stat_compare_means(comparisons =list(c("disease free","recurred")), method = 'wilcox.test', size=3.5) +
  facet_grid(gene ~ cohort) 



# Model Building : Elastic Net Regression 
control <- trainControl(method = "none", 
                        allowParallel   = TRUE,
                        classProbs = TRUE, 
                        savePredictions = "final",
                        verboseIter = TRUE) 
  
jointDfMod           = jointDf
colnames(jointDfMod) = gsub(" ", ".", colnames(jointDfMod))  
jointDfMod$Disease.Free.Status = factor(gsub(" ", ".", as.character(jointDfMod$Disease.Free.Status)))

# Training ELastic Net Regression model 
elastic_model_a <- caret::train(Disease.Free.Status ~ cohort + U2AF2 + RUVBL1 + HDGF + FABP4 + STMN1, 
                                data       = jointDfMod,
                                method     = "glmnet",
                                metric     = "ROC",
                                family     = "binomial",
                                trControl  = control,
                                trace      = FALSE)
elastic_model_a

rocinput_a = data.frame(recurred = predict(elastic_model_a, type="prob")[,2], 
                        obs = elastic_model_a$trainingData$.outcome) 
roc_a <- roc(obs ~ recurred, data = rocinput_a)

# tcga
elastic_model_t <- caret::train(Disease.Free.Status ~ U2AF2 + RUVBL1 + HDGF + FABP4 + STMN1, 
                                data       = jointDfMod[jointDfMod$cohort == "TCGA",],
                                method     = "glmnet",
                                metric     = "ROC",
                                family     = "binomial",
                                trControl  = control,
                                trace      = FALSE)
elastic_model_t

rocinput_t = data.frame(recurred = predict(elastic_model_t, type="prob")[,2], 
                        obs = elastic_model_t$trainingData$.outcome) 
roc_t <- roc(obs ~ recurred, data = rocinput_t)

# cancer cell
elastic_model_c <- caret::train(Disease.Free.Status ~ U2AF2 + RUVBL1 + HDGF + FABP4 + STMN1, 
                                data       = jointDfMod[jointDfMod$cohort == "Cancer Cell",],
                                method     = "glmnet",
                                metric     = "ROC",
                                family     = "binomial",
                                trControl  = control,
                                trace      = FALSE)
elastic_model_c

rocinput_c = data.frame(recurred = predict(elastic_model_c, type="prob")[,2], 
                        obs = elastic_model_c$trainingData$.outcome) 
roc_c <- roc(obs ~ recurred, data = rocinput_c)

roc.list = list("Both Cohorts" = roc_a,
                "TCGA" = roc_t,
                "Cancer Cell" = roc_c)

ci.list <- lapply(roc.list, ci.se, specificities = seq(0, 1, l = 25))

dat.ci.list <- lapply(ci.list, function(ciobj) 
  data.frame(x = as.numeric(rownames(ciobj)),
             lower = ciobj[, 1],
             upper = ciobj[, 3]))

rocPlot1 <- ggroc(roc.list) + 
              theme_minimal() + 
              geom_abline(slope=1, 
              intercept = 1, 
              linetype = "dashed", 
              alpha=0.7,
              color = "grey") + 
              coord_equal() + 
              xlab("Specificity") + 
              ylab("Sensitivity") +
              theme(plot.title = element_text(size=11))

aucRes = rbind(c("Both Cohorts", roc_a$auc[1]),
c("TCGA only", roc_t$auc[1]),
c("Cancer Cell only", roc_c$auc[1]))

aucRes = data.frame(aucRes)
aucRes[,2] = 100*round(as.numeric(aucRes[,2]),4)
colnames(aucRes) = c("Data Used for GLMNet", "AUC Estimate")

aucResStr = paste0(aucRes[1,1], " AUC=", aucRes[1,2], "\n",
aucRes[2,1], " AUC=", aucRes[2,2], "\n",
aucRes[3,1], " AUC=", aucRes[3,2])

rocPlot1 <- rocPlot1 + ggtitle(aucResStr)

```

# Introduction

Our primary goal in this analysis is to assist with the
response to the following reviewer comment:

```
The authors simply averaged the Z-scores of each gene equally
without any adjusted weighting or attempts to use more 
sophisticated regression or elastic net methods. The authors 
may want to test if these approaches improve the performance 
of their signature on the TCGA and Taylor datasets
```

In this report, we explore different modeling strategies
to incorporate the five variables of interest to predict 
disease free vs recurrence status.

# Datasets for Analysis
## Cancer Cell
The source file for the Cancer Cell cohort was `Logistic Regression Cancer Cell.csv`

## TCGA
The source file for the Cancer Cell cohort was `Logistic Regression TCGA.csv`

# Summary of Results
Below is a dynamic table displaying all joined sample results across
the two cohorts. Note that some values are missing and these
rows were not utilized in the model fitting analysis. 

This table can be filtered, sorted,
or exported to `csv` or `xlsx` format. 

```{r echo=FALSE, warning=FALSE}
myDT1 
```

After removal of missing data, we have the following 
counts per disease status category and cohort for model fitting:

```{r echo=FALSE, warning=FALSE}
myDT2
```

# Analysis Results
Here we summarize the overall results emphasizing classification of recurrence
status as a binary variable.

## Distributions
Looking at each gene and differences associated with recurrence, 
we find significant differences in both cohorts (Mann-Whitney test
shown below):

```{r echo=FALSE, fig.width=7.5,fig.height=9.5} 
  plotDist
```

## Elastic Net Regression
Using the `glmnet` package in R, we performed
of elastic net model fitting using all data (no cross validation).
This was done merging both cohorts together, and for TCGA and
Cancer Cell cohorts individually. Below we display the ROC/AUC
results of those model fits. 

```{r echo=FALSE, fig.width=7.5,fig.height=6.5} 
  rocPlot1
```

## Variable Importance
Below we show the variable importance results
for each of the three elastic net model fitting approaches.

### Both cohorts
```{r echo=FALSE}
  varImp(elastic_model_a)
```

### TCGA only
```{r echo=FALSE}
  varImp(elastic_model_t)
```

### Cancer Cell only
```{r echo=FALSE}
  varImp(elastic_model_c)
```

## Final Model Coefficients
Below we display the coefficients in the final
model for each input dataset

### Both cohorts
```{r echo=FALSE}
  coef(elastic_model_a$finalModel, elastic_model_a$bestTune$lambda)

```

### TCGA only
```{r echo=FALSE}
  coef(elastic_model_t$finalModel, elastic_model_t$bestTune$lambda)

```

### Cancer Cell only
```{r echo=FALSE}
  coef(elastic_model_c$finalModel, elastic_model_c$bestTune$lambda)
```

# End of Document